Running Group Networks
======================
Last update by Benjamin Chan (<benjamin.ks.chan@gmail.com>) on `r paste(Sys.time())` using `r R.version.string`.

At the Pre-holiday [fat ass](http://blog.oregonlive.com/runoregon/2009/10/fat_ass_runs_in_oregon_and_sw.html) in Forest Park, December 15, 2012, I said to [Josh Owen](https://www.facebook.com/joshowen) and [Pete Savage](https://www.facebook.com/pete.savage.56) that the fat ass was a conglomeration between the BananaSluggers and the [TrailFactor](http://www.facebook.com/TrailFactor) groups; and that there's a 50-80% overlap between the two groups. I even volunteered to create a network graph of the two groups. Here it is.


Getting Facebook network data
-----------------------------
Here are the steps to getting the data from Facebook into a format for R.
* Use the [netvizz](https://apps.facebook.com/netvizz/) Facebook app to create GDF files for both the BananaSluggers and the TrailFactor friend connections (note, netvizz required many attempts and patience; after a couple of days of attempts it finally produced the GDF files)
* Use [Gephi](https://gephi.org/) to open the GDF file for the BananaSluggers group
* Create a boolean column, `GroupBananaSlugger`, indicating BananaSlugger membership
* Export the revised graph in GEXF format
* Use [Gephi](https://gephi.org/) to open the GDF file for the TrailFactor group
* Create a boolean column, `GroupTrailFactor`, indicating TrailFactor membership
* Export the revised graph in GEXF format
* Start a new Gephi project
* First open the BananaSlugger GEXF graph
* Next, open the TrailFactor GEXF graph, but use the *Append Graph* option instead of the New Graph option
* Export the graph to GML format

There's probably a way to get around using Gephi but I'm not sure if igraph, or some other SNA package, can merge two graphs.


Load required packages.
```{r LoadPackages}
require(igraph, quietly=TRUE)
require(ggplot2, quietly=TRUE)
require(xtable, quietly=TRUE)
require(RColorBrewer, quietly=TRUE)
```

```{r ReadGraph}
G <- read.graph(file="BananaSluggersTrailFactor.gml", format="gml")
```
There are `r vcount(G)` nodes and `r ecount(G)` edges.

Create first name and initials vectors from names. I'll want to use these to label nodes when plotting the network.
```{r NodeLabels}
listName <- strsplit(V(G)$label, " ")
nameF <- sapply(listName, head, 1)
nameL <- sapply(listName, tail, 1)
label <- nameF
df <- data.frame(nameF, nameL)
```


Analysis questions
------------------
I'm interested in the following questions
* What is the magnitude of the overlap of BananaSluggers and TrailFactor?
* Who are the key people in the combined network?


Overlap of both groups
----------------------
Create logical vectors for group membership.
```{r GroupMembership}
grpBS <- as.logical(V(G)$GroupBananaSluggers)
grpBS[is.na(grpBS)] <- FALSE
grpTF <- as.logical(V(G)$GroupTrailFactor)
grpTF[is.na(grpTF)] <- FALSE
```
Create a factor for group membership and add to the data frame.
```{r CreateMembershipDataFrame}
group <- rep(NA, length(label))
group[ grpBS & !grpTF] <- "BS"
group[!grpBS &  grpTF] <- "TF"
group[ grpBS &  grpTF] <- "BS+TF"
group <- factor(group)
df <- data.frame(df, group)
```
Display the group membership totals.
```{r MembershipTable}
t <- table(df$group)
t
```
`r format(t["BS+TF"] / sum(t) * 100, digits=2)`% of the nodes are in both groups.


Centrality
----------
Here's a link to [Wikipedia](http://en.wikipedia.org/wiki/Centrality) for some background information on centrality.

Calculate **degree** centrality. This ends up not being too useful. Degree is really just a measure of how connected someone is. In Facebook terms, it's who has the most friends within the BS+TF network.
```{r Degree}
deg <- degree(G)
summary(deg)
```
The median degree, or number of friends, was `r median(deg)`. The highest number of friends a person in the BS+TF network has was `r max(deg)`.

Even though I don't want to focus on degree centrality, let's see who are the most connected people in the BS+TF network.
```{r HighDegree, results='asis'}
lim <- sort(deg, decreasing=TRUE)[round(vcount(G) * 0.05)]
top <- data.frame("Name"=V(G)$label, "Group"=group, "Degree"=deg)
top <- subset(top, deg >= lim)
top <- top[order(top$Degree, decreasing=TRUE),]
print(xtable(top, digits=0), type="html", include.rownames=FALSE)
```

Calculate **closeness** centrality. Closeness is a measure of how many steps are required to access every other node. It's a measure of how close a node is to all the action. A person with high closeness, however, doesn't necessarily have to have very many friends or be in between relationship.
```{r Closeness}
close <- closeness(G)
summary(close)
```
Again, I don't want to focus on closeness centrality since it's not really what I'm after in this analysis, so I won't say anything more about it.

Calculate **betweenness** centrality. Betweenness is a measure of how often a node is in the pathway between two other nodes. I.e., a person with high betweenness can be a key player in introducing a large group of friends to another large group of friends. Such a person doesn't necessarily have to have a large number of friends themselves. But they could be in a unique position of influence in the network.

Summarize the betweenness scores. 
```{r Betweenness}
btwn <- betweenness(G)
summary(btwn)
```
The absolute value of the scores don't mean much. But their relative values tell the story. List the top 5% of people with the highest betweeness scores.
```{r HighBetweenness, results='asis'}
lim <- sort(btwn, decreasing=TRUE)[round(vcount(G) * 0.05)]
top <- data.frame("Name"=V(G)$label, "Group"=group, "Betweenness"=btwn)
top <- subset(top, btwn >= lim)
top <- top[order(top$Betweenness, decreasing=TRUE),]
print(xtable(top, digits=0), type="html", include.rownames=FALSE)
```

Plot the association between degree centrality and betweenness centrality. See if there are any highly influential people (betweenness) who also have a high number of friends (degree).
```{r AssociationCentrality}
rsq <- format(cor(deg, btwn) ^2, digits=3)
cntrl <- data.frame(deg, btwn, close)
xlab <- mean(c(min(cntrl$deg), max(cntrl$deg)))
ylab <- mean(c(min(cntrl$btwn), max(cntrl$btwn)))
ggplot(cntrl, aes(x=deg, y=btwn)) +
  geom_jitter(alpha=1/2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x="Degree", y="Betweenness") +
  annotate("text", label=paste("R-sq =", rsq), x=xlab, y=log(ylab))
```


**NEED TO EDIT**
Visualize the network
---------------------
Label the nodes with high betweenness centrality.
```{r SubsetLabels}
topnodes <- nameF
topnodes[btwn < lim] <- NA
```
Scale the size of a node's plotting symbol according to its betweenness centrality score. Scaling is by percentile (95%, 98%, 99%).
```{r ScaleSymbol}
q <- quantile(btwn, probs=c(0, 0.95, 0.98, 0.99, 1))
size <- cut(btwn, breaks=q, include.lowest=TRUE, dig.lab=5)
summary(size)
V(G)$size <- unclass(size) + 1
# V(G)$size <- 1
```

Plot the network and color-code the groups.
```{r NetworkVisualization, fig.height=10, fig.width=10}
n <- nlevels(df$group)
palette <- brewer.pal(n, "Spectral")
V(G)$color[group == "BS"   ] <- palette[2]
V(G)$color[group ==    "TF"] <- palette[1]
V(G)$color[group == "BS+TF"] <- palette[3]
title <- paste("BananaSluggers + TrailFactor", "Facebook Network", sep="\n")
plot(G, 
     main=title,
     vertex.label=NA, 
     vertex.label.color="black", 
     vertex.label.dist=0.5, 
     vertex.label.family="sans", 
     vertex.frame.color="gray",
     edge.color="lightgray",
     edge.curved=TRUE,
     )
```

Whoa! That's a pretty intense network graph. Let's prune it a bit and only show the runners with high degree.
```{r PruneGraph}
p <- 75
cutoff <- quantile(deg, p / 100)
drop <- deg < cutoff
G <- delete.vertices(G, drop)
topnodes <- topnodes[!drop]
group <- group[!drop]
V(G)$size <- unclass(size[!drop]) + 1
```
Plot runners in the top `r 100 - p`% in terms of degree. I.e., nodes with degree > `r format(cutoff, digits=4)`.
```{r NetworkVisualizationPruned, fig.height=15, fig.width=15}
subtitle <- paste("Members with degree \u2265 ", format(cutoff, digits=4), " (", p, "th percentile)", sep="")
title <- paste(title, subtitle, sep="\n")
plot(G, 
     main=title,
     vertex.label=topnodes, 
     vertex.label.color="black", 
     vertex.label.dist=0.5, 
     vertex.label.family="sans", 
     vertex.frame.color="gray",
     edge.color="lightgray",
     edge.curved=T
     )
```
